---
title: "Identify contaminants with decontam and manually"
author: "Mari-Lee Odendaal (adapted by Beatrice Cornu Hewitt)"
date: "`r Sys.time()`"
output: html_document
---

The overall goal of this script is to identify the contaminants per niche using Decontam and our own manual filtering step. After careful exploration, the list of list of contaminants per niche are stored and later used to exclude the contaminant ASVs from the raw phyloseq object.

```{r setup, include=F, message=F, include=F}
# Define a function to check and install packages
install_if_missing <- function(packages) {
  new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(new_packages)) install.packages(new_packages)
}
# List of required packages
required_packages2 <- c(
  "tidyverse", "magrittr", "glue", "ggtext", "fs", "phyloseq", "microbiomer", "patchwork",
  "microbiome", "vegan", "microbiome", "decontam", "reshape2", "colorspace", "scales", "ggpubr", "ggvenn", "ggnewscale", 
  "RColorBrewer", "paletteer", "gridExtra", "ggdist"
)
# Install missing packages
install_if_missing(required_packages2)
# Load all required packages
lapply(required_packages2, library, character.only = TRUE)
```

# Theme for plots

```{r setup, include=F, message=F, include=F}
theme_set(theme_bw())
theme_update(axis.text.x = element_text(angle = 45, hjust=1),
             strip.text = element_text(colour = 'white'),
             strip.background =element_rect(fill="#2F4858", color = "#2F4858"))
```

# Functions
The functions used in this RMarkdown file are stored in "functions.R"
```{r}
source("functions.R")
```

## Input
As input data we will need the previously created raw phyloseq object.

```{r}
ps1 <- readRDS("../Output/Phyloseq/ps_raw.Rds")
ps_raw <- ps1
```

# Exploration
The library size (number of reads per sample) of the samples are ordered. The blanks have lower reads compared to the nasopharynx samples.

```{r library_size_per_sample_type, ow = '90%', h = 2.5, w = 5}
df <- meta_to_df(ps1)
df <- subset(df, niche != "NP?") %>% subset(., niche != "BB") %>% 
  subset(., niche != "BP") %>% subset(., niche != "UMD") %>% 
  subset(., niche != "ZMCD")
df <- df[order(df$after_DADA2_reads_nr),]
df$Index <- seq(nrow(df))
libsize <- ggplot(data = df, aes(x=Index, y=after_DADA2_reads_nr, color=niche)) + geom_point() + 
    scale_color_manual(values = c("burlywood3", "cadetblue3", 
                                  "coral4", "#FFC857", "darkolivegreen4")) 

ggsave("../Output/Microbiome_descriptives/Library_size_persample.png",libsize)
```

The overall composition of the NP samples looks like a typical nasopharyngeal microbiota: High abundances of Moraxella, Staphylococcus, Corynebacterium and Dolosigranulum. However, we also find the typical contaminant Pseudomonas and Caulobacteraceae high on the list. Interestingly, the patients have a high relative abundance of Streptococcus and Haemophilus.


```{r composition_NP, ow = '90%', h = 5, w = 6}
NP_composition <- create_ordered_bar(subset_samples(ps1, !is.na(population)), "NP", 15) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~population, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_NP.png", NP_composition)
ggsave("../Output/Microbiome_descriptives/Composition_plot_NP.svg", NP_composition)

```
The overall composition of the OP samples looks like a typical oropharyngeal microbiota: Streptococcus, Veilonella, Prevotella, Neisseria, etc.

```{r composition_OP, ow = '90%', h = 5, w = 6}
OP_composition <- create_ordered_bar(subset_samples(ps1, !is.na(population)), "OP", 15) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~population, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_OP.png", OP_composition)
ggsave("../Output/Microbiome_descriptives/Composition_plot_OP.svg", OP_composition)

```

The BD controls have high abundances of Pseudomonas, Caulobacteraceae, Bradyrhizobium, Burkholderia, etc.

```{r composition_BD, ow = '90%', h = 5, w = 5.5}
BD_controls_composition <- create_ordered_bar(ps1, "BD", 10) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~niche, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_BD_controls.png", BD_controls_composition)

```

The FB controls have high abundances of Pseudomonas Caulobacteraceae, Bradyrhizobium, etc.

```{r composition_FB, ow = '90%', h = 5, w = 4}
FB_controls_composition <- create_ordered_bar(ps1, "FB", 10) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~niche, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_FBcontrols.png", FB_controls_composition)
```

The Zymomocks have a stable composition.

```{r composition_ZMCB, ow = '90%', h = 5, w = 4}
ZMCB_controls_composition <- create_ordered_bar(ps1, "ZMCB", 10) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~niche, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_ZMCB_controls.png", ZMCB_controls_composition)
```

```{r composition_ZMCD, ow = '90%', h = 5, w = 4}
ZMCD_controls_composition <- create_ordered_bar(ps1, "ZMCD", 10) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~niche, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_ZMCD_controls.png", ZMCD_controls_composition)

```

The Zymomocks have a stable composition.

```{r composition_UMD, ow = '90%', h = 5, w = 4}
UMD_controls_composition <- create_ordered_bar(ps1, "UMD", 10) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~niche, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_UMD_controls.png", UMD_controls_composition)

```

# Cleaning phyloseq
## Exclusion of mocks
After checking the composition of the Zymo mocks, we remove them together with some other controls from our dataset

```{r}
ps1 <- subset_samples(ps1, niche != "ZMCB") %>% 
  subset_samples(., niche != "ZMCD") %>%  subset_samples(., niche != "BB") %>%
  subset_samples(., niche != "BP") %>% subset_samples(., niche != "NP?") %>%
  subset_samples(., qPCR != 0)
sample_data(ps1)$is_neg <- sample_data(ps1)$niche == "BD"

ps2 <- ps1
ps1 <- subset_samples(ps1, niche != "FB") 
df <- meta_to_df(ps1)
```

## Identifying contaminants using Decontam
We use the R-package "Decontam" to identify contaminants from our dataset stratified by niche (OP and NP). We identified a total of 128 contaminant in NP and 115 in OP using the combined method in decontam

```{r}
contaminants_np <- isContaminant(subset_samples(ps1, 
                                                niche == "NP" | niche == "BD"), 
                                 method="combined", conc="qPCR", neg="is_neg")

contaminants_np <- subset(contaminants_np, contaminant == TRUE)
dim(contaminants_np)[1] # 128 identified
```

```{r}
contaminants_op <- isContaminant(subset_samples(ps1, 
                                                niche == "OP" | niche == "BD"), 
                                 method="combined", conc="qPCR", neg="is_neg")

contaminants_op <- subset(contaminants_op, contaminant == TRUE)
dim(contaminants_op)[1] # 115
```

23 ASVs were identified as contaminants in both the NP and OP.

```{r}
con_both <- rownames(contaminants_op)[rownames(contaminants_op) %in% rownames(contaminants_np)]
con_both
```

We create a frequency plot of the top 30 identified contaminants. For most of these contaminant ASVs there seems to be a correlation between their relative abundance and the bacterial density.

```{r decontam_freq_np, ow = '90%', h = 15, w = 15}
contam_NP_freqplot <- plot_frequency(subset_samples(ps1, niche == "NP"), 
               head(rownames(contaminants_np), 30), 
               conc="qPCR") + xlab("Bacterial density (pg/ul)")

contam_NP_freqplot
```

```{r decontam_freq_op, ow = '90%', h = 15, w = 15}
contam_OP_freqplot <- plot_frequency(subset_samples(ps1, niche == "OP"), 
               head(rownames(contaminants_op), 30), 
               conc="qPCR") + xlab("Bacterial density (pg/ul)")
contam_OP_freqplot
```

The contaminants identified are shown in the table. These ASVs were also previously identified as the most abundant ASVs in the blanks.

```{r}
contaminants_op[which(contaminants_op$contaminant),]
contaminants_np[which(contaminants_np$contaminant),]
```

## Exploration of identified contaminants
Next, we convert the phyloseq to relative abundance, change the zero counts in a extremely low percentage (Lowest percentage divided by two)

```{r}
ps_ra <- subset_samples(ps1, niche == "NP" | niche == "OP") %>% to_RA()
```

All the identified contaminants found in the niches were screened across MiSeq runs and showed the profiles of a typical contaminant.

```{r decontam_samples_combined, ow = '90%', h = 8, w = 17}
df <- prune_taxa(con_both, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(con_both), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
#plot_contaminants(df)
```

```{r decontam_samples_OP, ow = '90%', h = 20, w = 20}
con <- rownames(contaminants_op)[!rownames(contaminants_op) %in% con_both][1:40]
df <- prune_taxa(con, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(con), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
#plot_contaminants(df)
```

```{r decontam_samples_OP_NP, ow = '90%', h = 20, w = 20}
#plot_contaminants(subset(df, niche == "NP"))
```

```{r decontam_samples_OP_OP, ow = '90%', h = 20, w = 20}
#plot_contaminants(subset(df, niche == "OP"))
```

```{r decontam_samples_NP, ow = '90%', h = 20, w = 20}
con <- rownames(contaminants_np)[!rownames(contaminants_np) %in% con_both][1:40]
df <- prune_taxa(con, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(con), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
#plot_contaminants(df)
```

```{r decontam_samples_NP_NP, ow = '90%', h = 20, w = 20}
#plot_contaminants(subset(df, niche == "NP"))
```

```{r decontam_samples_NP_OP, ow = '90%', h = 20, w = 20}
#plot_contaminants(subset(df, niche == "OP"))
```

## Identify additional layer of contaminants
Because we have observed an additional layer of contaminants, we implement a Kruskal-Wallis test to test which ASVs show discrepancies across the DNA isolation runs

```{r}
ps_np <- subset_samples(ps1, niche == "NP") %>% pres_abund_filter() %>% to_RA() %>% 
  pres_abund_filter(pres = 20, abund = 0.0001)
ps_s_n <- ps_np %>% pres_abund_filter() %>% to_RA() %>% 
  pres_abund_filter(pres = 30, abund = 0.0001)
ps_np <- prune_taxa(!taxa_names(ps_np) %in% taxa_names(ps_s_n) , ps_np)

ps_op <- subset_samples(ps1, niche == "OP") %>% pres_abund_filter() %>% to_RA() %>% 
  pres_abund_filter(pres = 20, abund = 0.0001)
ps_s_n <- ps_op %>% pres_abund_filter() %>% to_RA() %>% 
  pres_abund_filter(pres = 30, abund = 0.0001)
ps_op <- prune_taxa(!taxa_names(ps_op) %in% taxa_names(ps_s_n) , ps_op)
```

```{r}
df_np <- pivot_longer(ps_to_df(ps_np), cols = taxa_names(ps_np)[!taxa_names(ps_np) %in% c("1174_901_12_1422")], names_to = "ASV", 
                   values_to = "RA")

df_op <- pivot_longer(ps_to_df(ps_op), cols = taxa_names(ps_op), names_to = "ASV", 
                   values_to = "RA")
```

```{r}
s_np <- data_frame(ASV = taxa_names(ps_np)[!taxa_names(ps_np) %in% c("1174_901_12_1422")],fit = NA) %>% 
  mutate(map(ASV, ~ kruskal.test(RA ~ DI.Run, data = subset(df_np, ASV == .x)) %>%
               broom::tidy())) %>% unnest()

s_np$Contaminant <- ifelse(s_np$p.value < 0.001, TRUE, FALSE)
contaminants_op[!rownames(contaminants_op) %in% s_np[s_np$Contaminant,]$ASV,]
```

```{r}
s_op <- data_frame(ASV = taxa_names(ps_op),fit = NA) %>% 
  mutate(map(ASV, ~ kruskal.test(RA ~ DI.Run, data = subset(df_op, ASV == .x)) %>%
               broom::tidy())) %>% unnest()

s_op$Contaminant <- ifelse(s_op$p.value < 0.001, TRUE, FALSE)
contaminants_np[!rownames(contaminants_np) %in% s_op[s_op$Contaminant,]$ASV,]
```

```{r}
ls_con <- list(NP_decontam = rownames(contaminants_np),
               OP_decontam = rownames(contaminants_op),
               NP_additional = s_np[s_np$Contaminant,]$ASV,
               OP_additional = s_op[s_op$Contaminant,]$ASV)

all_names <- unique(c(ls_con$NP_decontam, ls_con$OP_decontam, 
                      ls_con$NP_additional, ls_con$OP_additional))
result_df <- data.frame(name = all_names)
result_df$NP_decontam <- result_df$name %in% ls_con$NP_decontam
result_df$OP_decontam <- result_df$name %in% ls_con$OP_decontam
result_df$NP_additional <- result_df$name %in% ls_con$NP_additional
result_df$OP_additional <- result_df$name %in% ls_con$OP_additional
```

```{r}
con <- c(rownames(contaminants_np), rownames(contaminants_op)) %>% unique()
con <- con[!con %in% c("Anaerococcus_129", "Anaerococcus_provencensis_172", "Dolosigranulum_pigrum_1",
          "Enterobacteriaceae_123", "Neisseriaceae_144", "Staphylococcus_2")]  
```

```{r additional_samples_NP, ow = '90%', h = 12, w = 20}
s_np <- s_np[!s_np$ASV %in% con,]
s_np <- s_np[s_np$Contaminant == TRUE,]

df <- prune_taxa(s_np$ASV, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(s_np$ASV), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
#plot_contaminants(df)
```

```{r additional_samples_NP_NP, ow = '90%', h = 12, w = 20}
#plot_contaminants(subset(df, niche == "NP"))
```

```{r additional_samples_NP_OP, ow = '90%', h = 12, w = 20}
#plot_contaminants(subset(df, niche == "OP"))
```

```{r additional_samples_OP, ow = '90%', h = 6, w = 15}
s_op <- s_op[!s_op$ASV %in% con,]
s_op <- s_op[s_op$Contaminant == TRUE,]

df <- prune_taxa(s_op$ASV, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(s_op$ASV), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
#plot_contaminants(df)
```

## Check "false positives"
After screening the contaminants, we found six ASVs that were labeled wrongly as a contaminant by Decontam or our manual screen. The figures demonstrating this further are created below.

```{r}
ncon <- c("Anaerococcus_129", "Anaerococcus_provencensis_172", "Dolosigranulum_pigrum_1",
          "Enterobacteriaceae_123", "Neisseriaceae_144", "Staphylococcus_2")
```

```{r ncon_freq_op, ow = '90%', h = 2.5, w = 16}
plot_frequency(subset_samples(ps1, niche == "OP"), ncon, conc="qPCR") + 
  xlab("Bacterial density (pg/ul)") + facet_grid(~taxa)
```

```{r ncon_freq_np, ow = '90%', h = 2.5, w = 16}
plot_frequency(subset_samples(ps1, niche == "NP"), ncon, conc="qPCR") + 
  xlab("Bacterial density (pg/ul)") + facet_grid(~taxa)
```

```{r ncon_across_isolation_run_op, ow = '90%', h = 2.5, w = 16}
df <- prune_taxa(ncon, ps_ra) %>% ps_to_df()
df <- pivot_longer(df, all_of(ncon), names_to = "ASV")
df$value <- df$value + unique(sort(df$value))[2]/2
df$ASV <- factor(df$ASV, levels=unique(df$ASV)) %>% format_OTU()
plot_contaminants(subset(df, niche == "OP"))  +
    scale_colour_manual(values = c("cadetblue4")) + 
  facet_wrap(~ASV, ncol = 8)
```

```{r ncon_across_isolation_run_np, ow = '90%', h = 2.5, w = 16}
plot_contaminants(subset(df, niche == "NP")) + 
  facet_wrap(~ASV, ncol = 8)
```

## Percentage contamination
Bases on the Decontam and manual screen we create one list with all the contaminants. Using this list, we calculate the percentage of contamination found in each sample.

```{r}
con_all <- result_df$name[!result_df$name %in% ncon] 
ps_con <- prune_taxa(con_all, ps2)
ps_con@sam_data$contaminants_reads <- colSums(ps_con@otu_table)
m_con <-  meta_to_df(ps_con)
m_con$contaminants_per <- m_con$contaminants_reads/m_con$after_DADA2_reads_nr
m_con <- m_con %>% 
  mutate(population = case_when(
    niche == "FB" ~ "LC",
    TRUE ~ population
  ))
```

```{r bacterial_density_across_MiSeq_run, ow = '90%', h = 3, w = 8}
plot_sample_data(m_con, x=m_con$niche, y=m_con$qPCR) + 
  labs(x = "Sample Type", y = "qPCR (pg/µl)") + 
  scale_y_continuous(trans='log10', breaks=trans_breaks('log10', function(x) 10^x),
                     labels=trans_format('log10', math_format(10^.x))) +
    geom_hline(aes(yintercept = 0.095),  colour="grey50", linetype="dotted") + 
  facet_grid(~Miseq.Run, scale = "fixed")
```

```{r contamination_across_MiSeq_run, ow = '90%', h = 3, w = 8}
plot_sample_data(m_con, x=m_con$niche, y=m_con$contaminants_per) + 
  labs(x = "Sample Type", y = "Contamination (%)") +
    geom_hline(aes(yintercept = 0.9),  colour="grey50", linetype="dotted") + 
  facet_grid(~Miseq.Run, scale = "fixed")
```

5 samples have more than 90% contamination

```{r}
t <- subset(m_con, contaminants_per > 0.9) %>% 
  subset(niche != "BD") %>% subset(niche != "FB")
m_con <- m_con %>% select(sample_id, contaminants_per)
#ps1 <- prune_samples(!sample_names(ps1) %in% t$sample_id, ps1)
```

## Exclusion of blanks
We then exclude all the blanks, which leads to a total of 248 nasopharyngeal samples in our dataset (47 blanks excluded)

```{r}
ps_raw
ps1 <- prune_samples(sample_data(ps1)$niche !="BD", ps1)
ps1 <- prune_taxa(taxa_sums(ps1) > 0, ps1)
stats <- tibble(step = character(),
                samples = numeric(),
                taxa = numeric(),
                reads = numeric())
stats <- ps_stats(ps1, stats, "Exclusion of controls")
```

## Remove full contaminant list
After checking the contaminants across isolation run, we found 6 identified contaminants that showed profiles of normal non contaminant ASVs, therefore they were removed from the list.

```{r}
con_all
```

Then, we subset the contaminant ASVs identified by Decontam and the manual identification step (n = 289)

```{r}
ps1
ps1 <- prune_taxa(!taxa_names(ps1) %in% con_all, ps1)
ps1 <- prune_taxa(taxa_sums(ps1) > 0, ps1)

stats <- ps_stats(ps1, stats, "Exclusion of contaminant ASVs")

```

## Exclusion low density samples
Next, we remove all the samples with low bacterial density (lower than 0.1 pg/ul). This includes 20 samples

```{r}
ps1
ps1 <- prune_samples(!sample_data(ps1)$qPCR < 0.095, ps1)
ps1 <- prune_taxa(taxa_sums(ps1) > 0, ps1)

stats <- ps_stats(ps1, stats, "Exclusion of samples with bacterial density <0.095 pg/uL")
```

## Exclusion of samples with less than 10000 reads
First we take a look at the distribution of # reads per sample

```{r distribution_reads, ow = '90%', h = 3, w = 6}
read_counts <- sample_sums(ps1) %>% enframe() %>% ggplot(aes(y = 1, x = value)) +
  ggdist::stat_halfeye( adjust = .5, height = .3, justification = -.3,
                        .width = 0, point_colour = NA) + 
  geom_point(size = 1.3, shape = 16, alpha = .2, 
             position = position_jitter(seed = 1, width = 0, height = .05)) +
  geom_boxplot(width = .12, alpha = 0.5, outlier.color = NA) +
  scale_x_log10(breaks = c(1000, 5000, 10000, 30000, 100000, 300000),
                labels = scales::comma) +
  coord_cartesian(ylim = c(0.9, NA), clip = "off") +
  labs(x = "Number of reads/sample", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  geom_vline(xintercept = 10000, linetype = "dashed", color = "red")
```

Next, we generate rarecurves.

```{r rarefaction_curves, ow = '90%', h = 3, w = 6}
otu_tab_rare <- ps1 %>% prune_samples(sample_sums(.) < 30000, .) %>% 
  otu_table() %>% as(., "matrix") %>% t

rare_curve_data <- rarecurve(otu_tab_rare, step = 10)

rare_curves <- map2_dfr(rare_curve_data, 
                        rownames(otu_tab_rare), ~enframe(.x) %>% 
                          mutate(sample_id = .y)) %>%
  mutate(name = str_remove(name, "N") %>% as.numeric) %>%
  ggplot(aes(x = name, y = value, group = sample_id)) +
    geom_line(alpha = 0.15) +
    scale_y_log10() +
    scale_x_continuous(breaks = c(0, 5000, 10000, 15000, 20000, 25000), 
                       labels = scales::comma) +
    geom_vline(aes(xintercept = 10000),  colour="grey50", linetype="dotted") +
    labs(x = "Sample size", y = "Number of species")

ggsave("../Output/Microbiome_descriptives/rarefaction_curves.svg", rare_curves)
```

```{r rare_curves_read_depth, ow = '90%', h = 3, w = 8, echo=FALSE}
readcounts_rarecurves <- grid.arrange(read_counts, rare_curves, ncol = 2)
ggsave("../Output/Microbiome_descriptives/readcounts_rarecurves.svg",readcounts_rarecurves)

```

Based on these figures, we chose a read cut-off at 10,000 reads. In the rarefaction curves, this was the point where the curves started flattening. 

```{r}
ps1
ps1 <- prune_samples(sample_sums(ps1) > 10000, ps1)
ps1 <- prune_samples(!is.na(ps1@sam_data$population), ps1)
ps1 <- prune_taxa(taxa_sums(ps1) > 0, ps1)

stats <- ps_stats(ps1, stats, "Exclusion of samples with reads < 10,000")
```

```{r sample_distribution, ow = '90%', h = 4, w = 5}
m <- meta_to_df(ps1)
t <- count(m, population, niche) %>% as.data.frame()
samples_perpopulation <- ggplot(t, aes(x = population, y = n, fill = niche)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("burlywood3", "cadetblue3")) +
  labs(x = "Population", y = "Number of samples", fill = "Niche") +
  geom_text(position = position_stack(vjust = 0.5), 
            aes(label = n, colour = niche), size = 5) +
  scale_colour_manual(values = c("burlywood4", "cadetblue4")) +
  guides(colour = "none")+ 
  theme(axis.text.x = element_text(size = 14),  # Adjust the size as needed
        axis.text.y = element_text(size = 14),
        axis.title = element_text(size = 14),
        legend.title =element_text(size = 14),
        legend.text = element_text(size=14))  # Adjust the size as needed
ggsave("../Output/Microbiome_descriptives/samples_perpopulation.svg", samples_perpopulation)
```

```{r composition_NP_cleaned, ow = '90%', h = 5, w = 6}
np_stacked_barplot_cleaned <- create_ordered_bar(ps1, "NP", 15) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~population, scale = "free")

ggsave("../Output/Microbiome_descriptives/np_stacked_barplot_cleaned.svg", np_stacked_barplot_cleaned)
```

# Bacterial density & Shannon index
Calculate the alpha diversity indices.

```{r}
m <- ps1 %>% meta_to_df()
m <- left_join(m, m_con)
```

```{r}
set.seed(77)
ps2 <- phyloseq::rarefy_even_depth(ps1, replace = T, trimOTUs = T)
adiv <- data.frame(
  "Observed" = phyloseq::estimate_richness(ps2, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(ps2, measures = "Shannon"))

adiv$sample_id <- gsub("\\.", "-", rownames(adiv))
m <- left_join(m, adiv)
m$qPCR_log <- log2(m$qPCR)
```

## Exclusion of duplicates
We check for duplicates/repeats and find 0

```{r}
m_np <- subset(m, niche == 'NP')
m_op <- subset(m, niche == 'OP')
duplicates <- m_np[duplicated(m_np$sample) | duplicated(m_np$sample, fromLast = TRUE), ]
duplicates <- m_op[duplicated(m_op$sample) | duplicated(m_op$sample, fromLast = TRUE), ]
```

```{r}
ps1@sam_data$final_read_nr <- colSums(ps1@otu_table)
```

```{r Shannon_index_across_MiSeq_run, ow = '90%', h = 4, w = 8}
plot_sample_data(m, x=m$population, y=m$Shannon) + 
  labs(x = "Population", y = "Shannon index") +
    geom_hline(aes(yintercept = 0.8),  colour="grey50", linetype="dotted") + 
  facet_grid(niche~Miseq.Run, scale = "fixed")
```

```{r Shannon_index_across_population, ow = '90%', h = 3, w = 4}
shannon_population <- plot_sample_data(m, x=m$population, y=m$Shannon) + 
  labs(x = "Population", y = "Shannon index") +
    geom_hline(aes(yintercept = 0.8),  colour="grey50", linetype="dotted") + 
  facet_grid(.~niche, scale = "fixed")

ggsave("../Output/Alpha_diversity/shannon_population.svg",shannon_population)
```

```{r Observed_species_across_MiSeq_run, ow = '90%', h = 4, w = 8}
plot_sample_data(m, x=m$population, y=m$Observed) + 
  labs(x = "Population", y = "Observed species") +
    geom_hline(aes(yintercept = 0.8),  colour="grey50", linetype="dotted") + 
  facet_grid(niche~Miseq.Run, scale = "fixed")
```

```{r Observed_species_across_population, ow = '90%', h = 3, w = 4}
observed_population <- plot_sample_data(m, x=m$population, y=m$Observed) + 
  labs(x = "Population", y = "Observed species") +
    geom_hline(aes(yintercept = 0.8),  colour="grey50", linetype="dotted") + 
  facet_grid(.~niche, scale = "fixed")

ggsave("../Output/Alpha_diversity/observed_population.svg", observed_population)

```

```{r Bacterial_density_across_MiSeq_run, ow = '90%', h = 4, w = 8}
#plot_sample_data(m, x=m$population, y=m$qPCR) + 
#  labs(x = "Population", y = "qPCR (pg/µl)") +
#  facet_grid(niche~Miseq.Run, scale = "fixed") + 
#  scale_y_continuous(trans='log10', breaks=trans_breaks('log10', function(x) 10^x),
#                     labels=trans_format('log10', math_format(10^.x)))
```

```{r Bacterial_density_across_population, ow = '90%', h = 3, w = 4}
bacterialdensity_population <- plot_sample_data(m, x=m$population, y=m$qPCR) + 
  labs(x = "Population", y = "qPCR (pg/µl)") +
  facet_grid(.~niche, scale = "fixed") + 
  scale_y_continuous(trans='log10', breaks=trans_breaks('log10', function(x) 10^x),
                     labels=trans_format('log10', math_format(10^.x)))

ggsave("../Output/Alpha_diversity/bacterialdensity_population.svg", bacterialdensity_population)
```

# Check removed samples

```{r}
m_raw <- meta_to_df(ps_raw) %>% subset(niche == "NP" | niche == "OP")
m_removed <- m_raw[!m_raw$sample_id %in% m$sample_id,]
table(m_removed$population, m_removed$niche)
table(m_removed$Miseq.Run)
table(m_removed$Miseq.Run)
```

# Save phyloseq objects
Finally, we save two phyloseq object:

```{r}
rownames(m) <- m$sample_id
m <- select(m, -c("sample_id"))
sample_data(ps1) <- m
```

```{r}
saveRDS(ps1, file = "../Output/Phyloseq/ps_complete.Rds")
```

# Cleaned composition

```{r composition_OP_cleaned, ow = '90%', h = 5, w = 6}
op_stacked_barplot_cleaned <- create_ordered_bar(ps1, "OP", 15) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~population, scale = "free")

ggsave("../Output/Microbiome_descriptives/Composition_plot_op_cleaned.svg", op_stacked_barplot_cleaned)
```

```{r}
ps_g <- ps1 %>% tax_glom(taxrank = "Genus") %>% to_RA() 
taxa_names(ps_g) <- ps_g@tax_table[,6]
ps_g <- ps_g %>% pres_abund_filter(pres = 10, abund = 0.01)
```

```{r Genus_composition_OP_cleaned, ow = '90%', h = 5, w = 6}
ps_RA <- ps_g %>% prune_samples(sample_data(ps_g)$niche == "OP", .) %>% to_RA()
otu_RA_m <- as(otu_table(ps_RA), "matrix")
  
bc <- vegdist(t(otu_RA_m), "bray")
hc <- hclust(bc, method = "average")

excl_cols <- c("sample_id", colnames(phyloseq::sample_data(ps_RA)))


df_topn <- ps_RA %>% get_topn(n = 15, residuals = TRUE) %>% 
  ps_to_df(sample_name = "sample_id") %>% 
  tidyr::pivot_longer(-dplyr::all_of(excl_cols),  
                      names_to = "OTU", values_to = "value") %>% 
  dplyr::mutate(OTU = .data$OTU %>% forcats::fct_inorder() %>% 
                  forcats::fct_rev()) %>% dplyr::arrange(.data$sample_id) %>% 
        dplyr::mutate(sample_id = forcats::fct_inorder(.data$sample_id))

df_topn %>%
    mutate(sample_id = fct_relevel(sample_id, hc$labels[hc$order])) %>%
    create_bar(df_topn = ., n = 15, ncol_legend = 1, name_legend = "ASV") +
    coord_flip() +
    theme(legend.position = "right") + 
    scale_fill_manual(values = c("#F0F0F0", paletteer_d("pals::stepped")[2:(15+1)])) +
    labs(fill = "Genus") + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) + 
  facet_wrap(~population, scale = "free")
```

# Sessioninfo
The session info is as follows:

```{r}
sessionInfo()
```
