---
title: "DA analysis"
author: "Beatrice Cornu Hewitt" 
date: "`r Sys.time()`"
---
# Packages
```{r}
library(phyloseq); library(dplyr); library(ANCOMBC); library(mia); library(DT); library(Maaslin2); library(microbiomer); library(patchwork);library(rcartocolor); library(ggplot2); library(scales); library(cartography); library(mia); library(rngtools); library(tidyverse); library(mia); library(car)

# Set the theme for ggplot
theme_set(theme_bw())
theme_update(
  axis.text.x = element_text(angle = 45, hjust = 1),
  strip.text = element_text(colour = 'white'),
  strip.background = element_rect(fill = "#2F4858", color = "#2F4858")
)
```
# Functions
```{r}
# Function to conduct MaAsLin2
do_maaslin2 <- function(ps, var, other_var, ref = NULL) { 
  #pre-processing
  sample_data(ps) <- sample_data(ps)[!is.na(sample_data(ps)[,var])]
  ps <- prune_taxa(taxa_sums(ps) > 0, ps)
  otu <- ps@otu_table %>% as.data.frame() %>% t() %>% as.data.frame()
  meta <- meta_to_df(ps)
  rownames(meta) <- meta$sample_id
  meta <- as.data.frame(meta)
  otu <- otu[rownames(otu) %in% rownames(meta),]
  #Maaslin2
  ms = Maaslin2(
    input_data = otu, 
    input_metadata = meta, 
    output = var, 
    fixed_effects = c(var, other_var), 
    normalization = "NONE",
    min_abundance = 0,
    min_prevalence = 0,
    plot_heatmap = FALSE,
    plot_scatter = FALSE,
    reference = ref)
  #Edit dataframe
  ms <- ms$results
  ms$var <- var
  
  ms <- ms[ms$metadata == var,]
  
  ms <- ms %>% mutate(conf.low = coef-1.96*stderr,
                      conf.high = coef+1.96*stderr,
                      label = ifelse(qval < 0.25, "*", NA),
                      label = ifelse(qval < 0.1, "**", label),
                      label = ifelse(qval < 0.05, "***", label)
                      )
  return(ms)
}
```
# Read in data
## CP only
```{r}
# Read in the phyloseq object which has the new metadata and the RF model predictions
ps_new_metadata_RFpreds <- readRDS("../Output/Phyloseq/ps_complete_new_metadata_RFpreds.rds")
# Remove any samples from pneumonia patients as not used in the analysis
ps_cp <- subset_samples(ps_new_metadata_RFpreds, population == "CP")
ps_cp <- prune_taxa(taxa_sums(ps_cp) > 0, ps_cp) # Also remove taxa that are not in the CP ps object

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_cp))$ASV)) # 17495
length(unique(as.data.frame(tax_table(ps_cp))$Genus)) # 960
```
## CP and GF 
```{r}
# CPs were matched 1:1 with the GFs (see other R notebook - "Matching_controls_goatfarmers.Rmd")
ps_cp_gf_matched <- readRDS("../Output/Phyloseq/ps_complete_matchedCPGF.rds")

# Count number of ASVs and genera in ps object
length(unique(as.data.frame(tax_table(ps_cp_gf_matched))$ASV)) # 7345
length(unique(as.data.frame(tax_table(ps_cp_gf_matched))$Genus)) # 761
```

## Check multicollinearity
```{r}
# Extract the sample data from the phyloseq object
sample_data_df <- data.frame(sample_data(ps_abs_cp_np_filt))
# Fit the linear model
lm_model <- lm(KRD_nGoatsWghtDist.3000m.sum ~ age + gender + sampling_season + smoked_ever +
                KRD_nCowsWghtDist.3000m.sum + KRD_nPoultryWghtDist.3000m.sum + 
                KRD_nPigsWghtDist.3000m.sum, data = sample_data_df)

# Calculate VIF values
vif_values <- vif(lm_model)

# Print the VIF values
print(vif_values)


# Correlation plot
# Extract the variables into a new data frame for simplicity
variables <- sample_data_df[, c("KRD_nPoultryWghtDist.3000m.sum", 
                                      "KRD_nCowsWghtDist.3000m.sum", 
                                      "KRD_nGoatsWghtDist.3000m.sum", 
                                      "KRD_nPigsWghtDist.3000m.sum")]


# Calculate the correlation matrix and p-values
corr_result <- psych::corr.test(variables)
corr_matrix <- corr_result$r    # Correlation matrix
p_matrix <- corr_result$p       # p-value matrix

# Convert correlation matrix and p-value matrix to long format
corr_long <- melt(corr_matrix)
p_long <- melt(p_matrix)

# Merge correlation and p-value matrices into one data frame
corr_long$p_value <- p_long$value

# Create labels based on significance levels
corr_long$label <- ifelse(corr_long$p_value < 0.001, "***",
                          ifelse(corr_long$p_value < 0.01, "**",
                                 ifelse(corr_long$p_value < 0.05, "*", "NS")))


# Only keep the upper triangle (excluding self-comparisons)
corr_long <- corr_long %>% 
  filter(as.numeric(Var1) < as.numeric(Var2))

# Plot the half-matrix heatmap with correlation coefficients and significance stars as labels
ggplot(corr_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#6D9EC1", high = "#E46726", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  geom_text(aes(label = paste0(round(value, 2), "\n", label)), color = "black", size = 4) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Correlation Matrix with Significance Stars")
```

# Data processing
## Filtering and conversion to relative abundance
### CP only
```{r}
# Apply the presence abundance filter (Subramanian et al., Nature, 2014: 0.1% relative abundance in at least 2 samples - we do this for the overall data at first - rare taxa can introduce noise so first apply this global filter. 
# Then convert the ASVs to relative abundances to account for different sequencing depths (normalization step)
ps_cp_RA <- ps_cp %>% 
  pres_abund_filter() %>% 
  to_RA() 
# A total of 2215 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 15280 ASVs excluded).

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_cp_RA))$ASV))
length(unique(as.data.frame(tax_table(ps_cp_RA))$Genus))
```
### CP and GF
```{r}
ps_cp_gf_RA <- ps_cp_gf_matched %>% 
  pres_abund_filter() %>% 
  to_RA() 
# A total of 1103 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 6242 ASVs excluded).

# Count number of genera left in ps object
length(unique(as.data.frame(tax_table(ps_cp_gf_RA))$ASV))
length(unique(as.data.frame(tax_table(ps_cp_gf_RA))$Genus))
```

## Convert to absolute abundances 
### CP only
```{r}
# Convert RA of ASVs by multiplying by the 16S qPCR measurements
cp_otu_tb <- otu_tab_to_df(ps_cp_RA)
cp_meta_tb <- meta_to_df(ps_cp_RA)
# Multiple RA by qPCR 
cp_otu_tb <- cp_otu_tb %>%
  mutate(across(-sample_id, ~.x*cp_meta_tb$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(cp_otu_tb) <- cp_otu_tb$sample_id
# Remove the first column, transpose and convert to df
cp_otu_tb <- cp_otu_tb[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
ps_abs_cp <- phyloseq(otu_table(cp_otu_tb, 
                             taxa_are_rows = T), 
                   tax_table(ps_cp_RA), 
                   sample_data(ps_cp_RA))
```
### CP and GF
```{r}
cp_gf_otu_tb <- otu_tab_to_df(ps_cp_gf_RA)
cp_gf_meta_tb <- meta_to_df(ps_cp_gf_RA)
# Multiple RA by qPCR 
cp_gf_otu_tb <- cp_gf_otu_tb %>%
  mutate(across(-sample_id, ~.x*cp_gf_meta_tb$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(cp_gf_otu_tb) <- cp_gf_otu_tb$sample_id
# Remove the first column, transpose and convert to df
cp_gf_otu_tb <- cp_gf_otu_tb[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
ps_abs_cp_gf <- phyloseq(otu_table(cp_gf_otu_tb, 
                             taxa_are_rows = T), 
                   tax_table(ps_cp_gf_RA), 
                   sample_data(ps_cp_gf_RA))
```
## Agglomerate data to genus level
### CP only
```{r}
ps_RA_genus_cp <- ps_abs_cp %>%
  tax_glom(taxrank = "Genus") %>% 
  to_RA() # 312 genera in the CP

# Change names to the genus-level names
taxa_names(ps_RA_genus_cp) <- ps_RA_genus_cp@tax_table[,6] 

# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
cp_genus_otu_tb <- otu_tab_to_df(ps_RA_genus_cp)
cp_genus_meta_tb <- meta_to_df(ps_RA_genus_cp)

# Multiple RA by qPCR 
cp_genus_otu_tb <- cp_genus_otu_tb %>%
  mutate(across(-sample_id, ~.x*cp_genus_meta_tb$qPCR)) %>% 
  as.data.frame()

# Make sample IDs the row names of the otu table
rownames(cp_genus_otu_tb) <- cp_genus_otu_tb$sample_id
# Remove the first column, transpose and convert to df
cp_genus_otu_tb <- cp_genus_otu_tb[,-1] %>% 
  t() %>%
  as.data.frame() 

# Create ps object with absolute abundance otu table
ps_genus_abs_cp <- phyloseq(otu_table(cp_genus_otu_tb, 
                             taxa_are_rows = T), 
                   tax_table(ps_RA_genus_cp), 
                   sample_data(ps_RA_genus_cp))
```
### CP and GF
```{r}
ps_RA_genus_cp_gf <- ps_abs_cp_gf %>%
  tax_glom(taxrank = "Genus") %>% 
  to_RA() # 247 genera

# Change names to the genus-level names
taxa_names(ps_RA_genus_cp_gf) <- ps_RA_genus_cp_gf@tax_table[,6] 

# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
cp_gf_genus_otu_tb <- otu_tab_to_df(ps_RA_genus_cp_gf)
cp_gf_genus_meta_tb <- meta_to_df(ps_RA_genus_cp_gf)

# Multiple RA by qPCR 
cp_gf_genus_otu_tb <- cp_gf_genus_otu_tb %>%
  mutate(across(-sample_id, ~.x*cp_gf_genus_meta_tb$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(cp_gf_genus_otu_tb) <- cp_gf_genus_otu_tb$sample_id

# Remove the first column, transpose and convert to df
cp_gf_genus_otu_tb <- cp_gf_genus_otu_tb[,-1] %>% 
  t() %>%
  as.data.frame() 

# Create ps object with absolute abundance otu table
ps_genus_abs_cp_gf <- phyloseq(otu_table(cp_gf_genus_otu_tb, 
                             taxa_are_rows = T), 
                   tax_table(ps_RA_genus_cp_gf), 
                   sample_data(ps_RA_genus_cp_gf))
```

## Filter per niche 
### CP only 
#### ASV level 
##### NP 
```{r}
samples_cp_np <- sample_data(ps_abs_cp)$niche == "NP" 
# Prune the phyloseq object to include only NP samples
ps_abs_cp_np <- prune_samples(samples_cp_np, ps_abs_cp)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_cp_np <- nrow(ps_abs_cp_np@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_abs_cp_np_filt <- ps_abs_cp_np %>%
  pres_abund_filter(pres = filter_cp_np, abund = 0.01)
# A total of 11 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 83.6 samples (n = 2201 ASVs excluded).

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_abs_cp_np_filt))$ASV)) # 11 ASVs
length(unique(as.data.frame(tax_table(ps_abs_cp_np_filt))$Genus)) # 9 genera
```
##### OP 
```{r}
# OROPHARYNX (OP)
samples_cp_op <- sample_data(ps_abs_cp)$niche == "OP" 
# Prune the phyloseq object to include only CP and OP samples
ps_abs_cp_op <- prune_samples(samples_cp_op, ps_abs_cp)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_cp_op <- nrow(ps_abs_cp_op@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_abs_cp_op_filt <- ps_abs_cp_op %>%
  pres_abund_filter(pres = filter_cp_op, abund = 0.01)
# A total of 40 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 97.8 samples (n = 2172 ASVs excluded).

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_abs_cp_op_filt))$ASV)) # 40 ASVs
length(unique(as.data.frame(tax_table(ps_abs_cp_op_filt))$Genus)) # 21 genera
```
#### Genus level 
##### NP
```{r}
samples_cp_np <- sample_data(ps_genus_abs_cp)$niche == "NP" 
# Prune the phyloseq object to include only CP and NP samples
ps_abs_genus_cp_np <- prune_samples(samples_cp_np, ps_genus_abs_cp)

# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_cp_np_genus <- nrow(ps_abs_genus_cp_np@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_abs_genus_cp_np_filt <- ps_abs_genus_cp_np %>%
  pres_abund_filter(pres = filter_cp_np_genus, abund = 0.01)
# A total of 10 GENERA were found to be present at or above a level of confident detection (1% relative abundance) in at least 83.6 samples (n = 279 ASVs excluded).

# Count number of genera left in ps object
length(unique(as.data.frame(tax_table(ps_abs_genus_cp_np_filt))$Genus)) # 10 genera
```
##### OP
```{r}
samples_cp_op <- sample_data(ps_genus_abs_cp)$niche == "OP" 
# Prune the phyloseq object to include only CP and OP samples
ps_abs_genus_cp_op <- prune_samples(samples_cp_op, ps_genus_abs_cp)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_cp_op <- nrow(ps_abs_genus_cp_op@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_abs_genus_cp_op_filt <- ps_abs_genus_cp_op %>%
  pres_abund_filter(pres = filter_cp_op, abund = 0.01)
# A total of 21 GENERA were found to be present at or above a level of confident detection (1% relative abundance) in at least 97.8 samples (n = 268 ASVs excluded).

# Check number of genera remaining
length(unique(as.data.frame(tax_table(ps_abs_genus_cp_op_filt))$Genus)) # 21 genera
```
### CP and GF
#### ASV level 
##### NP 
```{r}
samples_cp_gf_np <- sample_data(ps_abs_cp_gf)$niche == "NP" 
# Prune the phyloseq object to include only CP and NP samples
ps_np_cp_gf_abs <- prune_samples(samples_cp_gf_np, ps_abs_cp_gf)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_np_cp_gf <- nrow(ps_np_cp_gf_abs@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_np_cp_gf_abs_filt <- ps_np_cp_gf_abs %>%
  pres_abund_filter(pres = filter_np_cp_gf, abund = 0.01)
# A total of 13 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 16 samples (n = 1089 ASVs excluded).

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_np_cp_gf_abs_filt))$ASV)) # 13 ASVs
length(unique(as.data.frame(tax_table(ps_np_cp_gf_abs_filt))$Genus)) # 10 genera

```
##### OP 
```{r}
samples_cp_gf_op <- sample_data(ps_abs_cp_gf)$niche == "OP" 
# Prune the phyloseq object to include only CP and OP samples
ps_op_cp_gf_abs <- prune_samples(samples_cp_gf_op, ps_abs_cp_gf)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_op_cp_gf <- nrow(ps_op_cp_gf_abs@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_op_cp_gf_abs_filt <- ps_op_cp_gf_abs %>%
  pres_abund_filter(pres = filter_op_cp_gf, abund = 0.01)
# A total of 38 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 19.3 samples (n = 1064 ASVs excluded).

# Count number of ASVs and genera left in ps object
length(unique(as.data.frame(tax_table(ps_op_cp_gf_abs_filt))$ASV)) # 38 ASVs
length(unique(as.data.frame(tax_table(ps_op_cp_gf_abs_filt))$Genus)) # 20 genera
```
#### Genus level 
##### NP
```{r}
samples_cp_gf_np <- sample_data(ps_genus_abs_cp_gf)$niche == "NP"
# Prune the phyloseq object to include only CP and NP samples
ps_np_cp_gf_abs_genus <- prune_samples(samples_cp_gf_np, ps_genus_abs_cp_gf)
# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_np_cp_gf <- nrow(ps_np_cp_gf_abs_genus@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_np_cp_gf_abs_genus_filt <- ps_np_cp_gf_abs_genus %>%
  pres_abund_filter(pres = filter_np_cp_gf, abund = 0.01)

# A total of 12 GENERA were found to be present at or above a level of confident detection (1% relative abundance) in at least 16 samples (n = 214 ASVs excluded).

# Check number of genera remaining
length(unique(as.data.frame(tax_table(ps_np_cp_gf_abs_genus_filt))$Genus)) # 12 genera

```
##### OP
```{r}
samples_cp_gf_op <- sample_data(ps_genus_abs_cp_gf)$niche == "OP" 
# Prune the phyloseq object to include only CP and OP samples
ps_op_cp_gf_abs_genus <- prune_samples(samples_cp_gf_op, ps_genus_abs_cp_gf)

# Set up a presence abundance filter
# Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_op_cp_gf <- nrow(ps_op_cp_gf_abs_genus@sam_data) * 0.1 # 10% of total samples
# Apply the filter
ps_op_cp_gf_abs_genus_filt <- ps_op_cp_gf_abs_genus %>%
  pres_abund_filter(pres = filter_op_cp_gf, abund = 0.01)
# A total of 22 GENERA were found to be present at or above a level of confident detection (1% relative abundance) in at least 19.3 samples (n = 204 ASVs excluded).

# Check number of genera remaining
length(unique(as.data.frame(tax_table(ps_op_cp_gf_abs_genus_filt))$Genus)) # 22 genera
```

# Check metadata variables

```{r}
# MaAsLin fits a generalized linear model (GLM) for each feature (like microbial abundance) and metadata variable while adjusting for confounders and covariates. Here are the key assumptions and considerations for MaAsLin:
# 1. Independent Observations (e.g., no temporal or spatial autocorrelation).
# 2. Linearity (for linear models) - MaAsLin primarily models linear relationships between predictors (metadata) and outcomes (e.g., microbial abundance). Non-linear relationships between the metadata and microbial features may not be well captured.
# 3. Multicollinearity - metadata variables are not highly collinear with one another. 

# However MaAsLin can handle non-linear associations
# Select some example ASVs to plot
rownames(ps_abs_cp_np_filt@otu_table)
asvs_to_plot <- c("Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3")  # Replace with actual ASV names
metadata_vars <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", 
                   "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum", 
                   "ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", "meca_RF_preds")

# Extract metadata and ASV abundance data from phyloseq object
metadata <- sample_data(ps_abs_cp_np_filt) %>% data.frame()
asv_data <- otu_table(ps_abs_cp_np_filt) %>% as.data.frame()

metadata$KRD_nGoatsWghtDist.3000m.sum
# Convert ASV data to long format and combine with metadata
asv_data_long <- asv_data %>% t  %>% as.data.frame()
rownames(asv_data_long) <- asv_data_long$VGO3_ID

# Merge metadata with ASV data
merged_data <- cbind(asv_data_long, metadata, by = "VGO3_ID")

# Filter the merged data to only include the selected ASVs and metadata variables
plot_data <- merged_data %>%
  select(c("VGO3_ID", asvs_to_plot, metadata_vars))

# Scatterplots
ggplot(plot_data, aes(x = KRD_nGoatsWghtDist.3000m.sum, y = Dolosigranulum_pigrum_1)) +
  geom_point() +
  labs(x = "E.coli RF Predictions", y = "Dolosigranulum_pigrum_1 Abundance", 
       title = "Scatter Plot of KRD_nGoatsWghtDist.3000m.sum vs Dolosigranulum_pigrum_1") +
  theme_minimal()


# Histogram of ecoli_RF_preds
ggplot(plot_data, aes(x = ecoli_RF_preds)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "E.coli RF Predictions", y = "Frequency", 
       title = "Histogram of E.coli RF Predictions") +
  theme_minimal()

ggplot(plot_data, aes(x = KRD_nGoatsWghtDist.3000m.sum)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "E.coli RF Predictions", y = "Frequency", 
       title = "Histogram of KRD_nGoatsWghtDist.3000m.sum") +
  theme_minimal()

ggplot(plot_data, aes(x = KRD_nCowsWghtDist.3000m.sum)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "E.coli RF Predictions", y = "Frequency", 
       title = "Histogram of KRD_nCowsWghtDist.3000m.sum") +
  theme_minimal()

ggplot(plot_data, aes(x = KRD_nPigsWghtDist.3000m.sum)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "E.coli RF Predictions", y = "Frequency", 
       title = "Histogram of KRD_nPigsWghtDist.3000m.sum") +
  theme_minimal()



plot_data$KRD_nPigsWghtDist.3000m.sum


```


# MaAsLin2
## CP only
### ASV level 
#### NP niche
```{r}
# UNIVARIABLE for livestock species
# Define the variables to be analyzed
variables <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum",
               "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum",
               "ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds")

maaslin_results <- list()

output_file <- "../Output/DA_analysis/MaAsLin/CP_NP/"
# Loop and save each result to a CSV file
for (var in variables) {
  result <- do_maaslin2(ps = ps_abs_cp_np_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
                        arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0(output_file, "ASV_level_maaslin_cp_np_univar_", var, ".csv"))
}


# MULTIVARIABLE for livestock species
# Run MaAsLin for the (distance-based) livestock exposure variables (with adjustment for the other livestock variables)
# Goat farm exposure as main variable
Goat_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt, 
                               var = "KRD_nGoatsWghtDist.3000m.sum",  # Including all livestock variables together
                               other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                               ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
                               arrange(desc(coef))
# Save results as csv
write.csv(Goat_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_goatfarmexposure.csv"))

# Cattle farm exposure as main variable
Cow_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt, 
                               var = "KRD_nCowsWghtDist.3000m.sum",  # Including all livestock variables together
                               other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                               ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
                               arrange(desc(coef))
# Save results as csv
write.csv(Cow_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_cattlefarmexposure.csv"))

# Poultry farm exposure as main variable
Poultry_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt, 
                               var = "KRD_nPoultryWghtDist.3000m.sum",  # Including all livestock variables together
                               other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                               ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
                               arrange(desc(coef))
# Save results as csv
write.csv(Poultry_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_poultryfarmexposure.csv"))

# Pig farm exposure as main variable=
Pig_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt, 
                               var = "KRD_nPigsWghtDist.3000m.sum",  # Including all livestock variables together
                               other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nGoatsWghtDist.3000m.sum"),  # Adjusting for these variables
                               ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
                               arrange(desc(coef))
# Save results as csv
write.csv(Pig_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_pigfarmexposure.csv"))


# Loop for individual models for modelled exposures: ecoli, staph, tetw, and meca
RF_modelled_variables <- c("ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds")
# Create an empty list to store the ordered results
maaslin_results <- list()

# Loop and save each result to a CSV file
for (var in RF_modelled_variables) {
  result <- do_maaslin2(ps = ps_abs_cp_np_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
                        arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_", var, ".csv"))
}


```
<!-- #### NP niche - 2nd try with scaled predictors -->
<!-- ```{r} -->
<!-- # Scaling of predictor variables  -->

<!-- # Perform Z-score scaling on the continuous variables in your dataset -->
<!-- ps_abs_cp_np_filt_scaled <- ps_abs_cp_np_filt -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$age <- scale(sample_data(ps_abs_cp_np_filt_scaled)$age) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$KRD_nGoatsWghtDist.3000m.sum <- scale(sample_data(ps_abs_cp_np_filt_scaled)$KRD_nGoatsWghtDist.3000m.sum) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$KRD_nCowsWghtDist.3000m.sum <- scale(sample_data(ps_abs_cp_np_filt_scaled)$KRD_nCowsWghtDist.3000m.sum) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$KRD_nPoultryWghtDist.3000m.sum <- scale(sample_data(ps_abs_cp_np_filt_scaled)$KRD_nPoultryWghtDist.3000m.sum) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$KRD_nPigsWghtDist.3000m.sum <- scale(sample_data(ps_abs_cp_np_filt_scaled)$KRD_nPigsWghtDist.3000m.sum) -->

<!-- # If the RF modeled variables are continuous, scale them as well -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$ecoli_RF_preds <- scale(sample_data(ps_abs_cp_np_filt_scaled)$ecoli_RF_preds) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$staph_RF_preds <- scale(sample_data(ps_abs_cp_np_filt_scaled)$staph_RF_preds) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$tetw_RF_preds <- scale(sample_data(ps_abs_cp_np_filt_scaled)$tetw_RF_preds) -->
<!-- sample_data(ps_abs_cp_np_filt_scaled)$meca_RF_preds <- scale(sample_data(ps_abs_cp_np_filt_scaled)$meca_RF_preds) -->

<!-- # UNIVARIABLE for livestock species -->
<!-- # Define the variables to be analyzed -->
<!-- variables <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", -->
<!--                "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum", -->
<!--                "ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds") -->

<!-- maaslin_results <- list() -->

<!-- output_file <- "../Output/DA_analysis/MaAsLin/CP_NP/" -->
<!-- # Loop and save each result to a CSV file -->
<!-- for (var in variables) { -->
<!--   result <- do_maaslin2(ps = ps_abs_cp_np_filt_scaled, -->
<!--                         var = var, -->
<!--                         other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables -->
<!--                         ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories -->
<!--                         arrange(desc(coef)) -->
<!--   # Save to a CSV file -->
<!--   write.csv(result, paste0(output_file, "ASV_level_maaslin_cp_np_univar_scaled", var, ".csv")) -->
<!-- } -->


<!-- # MULTIVARIABLE for livestock species -->
<!-- # Run MaAsLin for the (distance-based) livestock exposure variables (with adjustment for the other livestock variables) -->
<!-- # Goat farm exposure as main variable -->
<!-- Goat_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt_scaled,  -->
<!--                                var = "KRD_nGoatsWghtDist.3000m.sum",  # Including all livestock variables together -->
<!--                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables -->
<!--                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories -->
<!--                                arrange(desc(coef)) -->
<!-- # Save results as csv -->
<!-- write.csv(Goat_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_goatfarmexposure_scaled.csv")) -->

<!-- # Cattle farm exposure as main variable -->
<!-- Cow_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt_scaled,  -->
<!--                                var = "KRD_nCowsWghtDist.3000m.sum",  # Including all livestock variables together -->
<!--                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables -->
<!--                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories -->
<!--                                arrange(desc(coef)) -->
<!-- # Save results as csv -->
<!-- write.csv(Cow_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_cattlefarmexposure_scaled.csv")) -->

<!-- # Poultry farm exposure as main variable -->
<!-- Poultry_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt,  -->
<!--                                var = "KRD_nPoultryWghtDist.3000m.sum",  # Including all livestock variables together -->
<!--                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables -->
<!--                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories -->
<!--                                arrange(desc(coef)) -->
<!-- # Save results as csv -->
<!-- write.csv(Poultry_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_poultryfarmexposure.csv")) -->

<!-- # Pig farm exposure as main variable= -->
<!-- Pig_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_np_filt,  -->
<!--                                var = "KRD_nPigsWghtDist.3000m.sum",  # Including all livestock variables together -->
<!--                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nGoatsWghtDist.3000m.sum"),  # Adjusting for these variables -->
<!--                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories -->
<!--                                arrange(desc(coef)) -->
<!-- # Save results as csv -->
<!-- write.csv(Pig_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_pigfarmexposure.csv")) -->


<!-- # Loop for individual models for modelled exposures: ecoli, staph, tetw, and meca -->
<!-- RF_modelled_variables <- c("ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds") -->
<!-- # Create an empty list to store the ordered results -->
<!-- maaslin_results <- list() -->

<!-- # Loop and save each result to a CSV file -->
<!-- for (var in RF_modelled_variables) { -->
<!--   result <- do_maaslin2(ps = ps_abs_cp_np_filt, -->
<!--                         var = var, -->
<!--                         other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables -->
<!--                         ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories -->
<!--                         arrange(desc(coef)) -->
<!--   # Save to a CSV file -->
<!--   write.csv(result, paste0("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/", "ASV_level_maaslin_cp_np_", var, ".csv")) -->
<!-- } -->


<!-- ``` -->

#### OP niche
```{r}
# UNIVARIABLE for livestock species
# Define the variables to be analyzed
variables <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum",
               "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum",
               "ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds" )

# Create an empty list to store the ordered results
maaslin_results <- list()
output_folder <- "../Output/DA_analysis/MaAsLin/CP_OP/"

# Loop and save each result to a CSV file
for (var in variables) {
  result <- do_maaslin2(ps = ps_abs_cp_op_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0(output_folder, "maaslin_cp_OP_asvlevel_", var, "_ord.csv"))
}

# MULTIVARIABLE for livestock species
# Run MaAsLin for the (distance-based) livestock exposure variables (with adjustment for the other livestock variables)
# Goat farm exposure as main variable
Goat_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_op_filt, 
                                 var = "KRD_nGoatsWghtDist.3000m.sum",  # Including all livestock variables together
                                 other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                 ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Goat_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/", "ASV_level_maaslin_cp_op_goatfarmexposure.csv"))

# Cattle farm exposure as main variable
Cow_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_op_filt, 
                                var = "KRD_nCowsWghtDist.3000m.sum",  # Including all livestock variables together
                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Cow_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/", "ASV_level_maaslin_cp_op_cattlefarmexposure.csv"))

# Poultry farm exposure as main variable
Poultry_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_op_filt, 
                                    var = "KRD_nPoultryWghtDist.3000m.sum",  # Including all livestock variables together
                                    other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                    ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Poultry_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/", "ASV_level_maaslin_cp_op_poultryfarmexposure.csv"))

# Pig farm exposure as main variable=
Pig_expo_maaslin <- do_maaslin2(ps = ps_abs_cp_op_filt, 
                                var = "KRD_nPigsWghtDist.3000m.sum",  # Including all livestock variables together
                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nGoatsWghtDist.3000m.sum"),  # Adjusting for these variables
                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Pig_expo_maaslin, paste0("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/", "ASV_level_maaslin_cp_op_pigfarmexposure.csv"))


# Loop for individual models for modelled exposures: ecoli, staph, tetw, and meca
RF_modelled_variables <- c("ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds")
# Create an empty list to store the ordered results
maaslin_results <- list()

# Loop and save each result to a CSV file
for (var in RF_modelled_variables) {
  result <- do_maaslin2(ps = ps_abs_cp_op_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/", "ASV_level_maaslin_cp_op_", var, ".csv"))
}
```
### Genus level 
#### NP niche
```{r}
# UNIVARIABLE for livestock species
# Define the variables to be analyzed
variables <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum",
               "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum",
               "ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds" )

# Create an empty list to store the ordered results
maaslin_results <- list()
output_folder <- "../Output/DA_analysis/MaAsLin/CP_NP/"

# Loop and save each result to a CSV file
for (var in variables) {
  result <- do_maaslin2(ps = ps_abs_genus_cp_np_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0(output_folder, "maaslin_cp_np_genuslevel_", var, "_ord.csv"))
}

# MULTIVARIABLE for livestock species
# Run MaAsLin for the (distance-based) livestock exposure variables (with adjustment for the other livestock variables)
# Goat farm exposure as main variable
Goat_expo_maaslin_genus <- do_maaslin2(ps = ps_abs_genus_cp_np_filt, 
                                 var = "KRD_nGoatsWghtDist.3000m.sum",  # Including all livestock variables together
                                 other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                 ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Goat_expo_maaslin_genus, paste0("../Output/DA_analysis/MaAsLin/CP_NP/genus_level/", "Genus_level_maaslin_cp_np_goatfarmexposure.csv"))

# Cattle farm exposure as main variable
Cow_expo_maaslin_genus <- do_maaslin2(ps = ps_abs_genus_cp_np_filt, 
                                var = "KRD_nCowsWghtDist.3000m.sum",  # Including all livestock variables together
                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Cow_expo_maaslin_genus, paste0("../Output/DA_analysis/MaAsLin/CP_NP/genus_level/", "Genus_level_maaslin_cp_np_cattlefarmexposure.csv"))

# Poultry farm exposure as main variable
Poultry_expo_maaslin_genus <- do_maaslin2(ps = ps_abs_genus_cp_np_filt, 
                                    var = "KRD_nPoultryWghtDist.3000m.sum",  # Including all livestock variables together
                                    other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                    ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Poultry_expo_maaslin_genus, paste0("../Output/DA_analysis/MaAsLin/CP_NP/genus_level/", "Genus_level_maaslin_cp_np_poultryfarmexposure.csv"))

# Pig farm exposure as main variable=
Pig_expo_maaslin_genus <- do_maaslin2(ps = ps_abs_genus_cp_np_filt, 
                                var = "KRD_nPigsWghtDist.3000m.sum",  # Including all livestock variables together
                                other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nGoatsWghtDist.3000m.sum"),  # Adjusting for these variables
                                ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Pig_expo_maaslin_genus, paste0("../Output/DA_analysis/MaAsLin/CP_NP/genus_level/", "Genus_level_maaslin_cp_np__pigfarmexposure.csv"))


# Loop for individual models for modelled exposures: ecoli, staph, tetw, and meca
RF_modelled_variables <- c("ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds")

# Loop and save each result to a CSV file
for (var in RF_modelled_variables) {
  result <- do_maaslin2(ps = ps_abs_genus_cp_np_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0("../Output/DA_analysis/MaAsLin/CP_NP/Genus_level/", "Genus_level_maaslin_cp_np_", var, ".csv"))
}

```

#### OP niche
```{r}
# UNIVARIABLE for livestock species
# Define the variables to be analyzed
variables <- c("KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum",
               "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum",
               "ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds" )

# Create an empty list to store the ordered results
maaslin_results <- list()
output_folder <- "../Output/DA_analysis/MaAsLin/CP_OP/"

# Loop and save each result to a CSV file
for (var in variables) {
  result <- do_maaslin2(ps = ps_abs_genus_cp_op_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0(output_folder, "maaslin_cp_op_genuslevel_", var, "_ord.csv"))
}

# MULTIVARIABLE for livestock species
# Run MaAsLin for the (distance-based) livestock exposure variables (with adjustment for the other livestock variables)
# Goat farm exposure as main variable
Goat_expo_maaslin_genus_cp_op <- do_maaslin2(ps = ps_abs_genus_cp_op_filt, 
                                       var = "KRD_nGoatsWghtDist.3000m.sum",  # Including all livestock variables together
                                       other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                       ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Goat_expo_maaslin_genus_cp_op, paste0("../Output/DA_analysis/MaAsLin/CP_OP/genus_level/", "Genus_level_maaslin_cp_op_goatfarmexposure.csv"))

# Cattle farm exposure as main variable
Cow_expo_maaslin_genus_cp_op <- do_maaslin2(ps = ps_abs_genus_cp_op_filt, 
                                      var = "KRD_nCowsWghtDist.3000m.sum",  # Including all livestock variables together
                                      other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                      ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Cow_expo_maaslin_genus_cp_op, paste0("../Output/DA_analysis/MaAsLin/CP_OP/genus_level/", "Genus_level_maaslin_cp_op_cattlefarmexposure.csv"))

# Poultry farm exposure as main variable
Poultry_expo_maaslin_genus_cp_op <- do_maaslin2(ps = ps_abs_genus_cp_op_filt, 
                                          var = "KRD_nPoultryWghtDist.3000m.sum",  # Including all livestock variables together
                                          other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nGoatsWghtDist.3000m.sum", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPigsWghtDist.3000m.sum"),  # Adjusting for these variables
                                          ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Poultry_expo_maaslin_genus_cp_op, paste0("../Output/DA_analysis/MaAsLin/CP_OP/genus_level/", "Genus_level_maaslin_cp_op_poultryfarmexposure.csv"))

# Pig farm exposure as main variable=
Pig_expo_maaslin_genus_cp_op <- do_maaslin2(ps = ps_abs_genus_cp_op_filt, 
                                      var = "KRD_nPigsWghtDist.3000m.sum",  # Including all livestock variables together
                                      other_var = c("age", "gender", "sampling_season", "smoked_ever", "KRD_nCowsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", "KRD_nGoatsWghtDist.3000m.sum"),  # Adjusting for these variables
                                      ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>%  # Set reference categories
  arrange(desc(coef))
# Save results as csv
write.csv(Pig_expo_maaslin_genus_cp_op, paste0("../Output/DA_analysis/MaAsLin/CP_OP/genus_level/", "Genus_level_maaslin_cp_op_pigfarmexposure.csv"))


# Loop for individual models for modelled exposures: ecoli, staph, tetw, and meca
RF_modelled_variables <- c("ecoli_RF_preds","staph_RF_preds","tetw_RF_preds","meca_RF_preds")

# Loop and save each result to a CSV file
for (var in RF_modelled_variables) {
  result <- do_maaslin2(ps = ps_abs_genus_cp_op_filt,
                        var = var,
                        other_var = c("age", "gender", "sampling_season", "smoked_ever"), # adjust for these variables
                        ref = c("gender,0", "sampling_season,Summer", "smoked_ever,0")) %>% # Set reference categories
    arrange(desc(coef))
  # Save to a CSV file
  write.csv(result, paste0("../Output/DA_analysis/MaAsLin/CP_OP/Genus_level/", "Genus_level_maaslin_cp_op_", var, ".csv"))
}

```
## CP and GF
### ASV level 
#### NP niche
```{r}
# Run Maaslin2 analysis with CP as reference category for population
maaslin_cp_gf_np <- do_maaslin2(ps = ps_np_cp_gf_abs_filt, 
                             var = "population", 
                             other_var = c("age", "gender", "sampling_season", "smoked_ever"), 
                             ref = c("population,CP", "gender,0", "sampling_season,Summer", "smoked_ever,0"))

# Reorder ASVs based on effect size
maaslin_cp_gf_np_ord <- maaslin_cp_gf_np %>%
  arrange(desc(coef))

write.csv(maaslin_cp_gf_np_ord, "../Output/DA_analysis/MaAsLin/CP_GF_NP/ASV_level/ASV_level_maaslin_cp_gf_np.csv")
```
#### OP niche
```{r}
# Run Maaslin2 analysis with CP as reference category for population
maaslin_cp_gf_op <- do_maaslin2(ps = ps_op_cp_gf_abs_filt, 
                             var = "population", 
                             other_var = c( "age", "gender", "sampling_season", "smoked_ever"), 
                             ref = c("population,CP", "gender,0", "sampling_season,Summer", "smoked_ever,0"))

# Reorder ASVs based on effect size
maaslin_cp_gf_op_ord <- maaslin_cp_gf_op %>%
  arrange(desc(coef))

write.csv(maaslin_cp_gf_op_ord, "../Output/DA_analysis/MaAsLin/CP_GF_OP/ASV_level/ASV_level_maaslin_cp_gf_op.csv")
```

### Genus level 
#### NP niche
```{r}
# Run Maaslin2 analysis with CP as reference category for population
maaslin_cp_gf_genus_np <- do_maaslin2(ps = ps_np_cp_gf_abs_genus_filt, 
                             var = "population", 
                             other_var = c("age", "gender", "sampling_season", "smoked_ever"), 
                             ref = c("population,CP", "gender,0", "sampling_season,Summer", "smoked_ever,0"))

# Reorder ASVs based on effect size
maaslin_cp_gf_genus_np_ord <- maaslin_cp_gf_genus_np %>%
  arrange(desc(coef))

write.csv(maaslin_cp_gf_genus_np_ord, "../Output/DA_analysis/MaAsLin/CP_GF_NP/Genus_level/maaslin_cp_gf_genus_np_ord.csv")
```
#### OP niche
```{r}
# Run Maaslin2 analysis with CP as reference category for population
maaslin_cp_gf_genus_op <- do_maaslin2(ps = ps_op_cp_gf_abs_genus_filt, 
                                      var = "population", 
                                      other_var = c("age", "gender", "sampling_season", "smoked_ever"), 
                                      ref = c("population,CP", "gender,0", "sampling_season,Summer", "smoked_ever,0"))

# Reorder ASVs based on effect size
maaslin_cp_gf_genus_op_ord <- maaslin_cp_gf_genus_op %>%
  arrange(desc(coef))

write.csv(maaslin_cp_gf_genus_op_ord, "../Output/DA_analysis/MaAsLin/CP_GF_OP/Genus_level/maaslin_cp_gf_genus_op_ord.csv")
```

# Heatmaps
## CP only
### ASV level 
#### NP (univariable species)
```{r}
# Define the directory
np_folder <- "../Output/DA_analysis/MaAsLin/CP_NP/"

# Read NP results
np_files_cp_asvlevel_univar <- list.files(np_folder, pattern = "ASV_level_maaslin", full.names = TRUE)
np_results_cp_asvlevel_univar <- lapply(np_files_cp_asvlevel_univar, read.csv)
np_results_cp_asvlevel_univar <- bind_rows(np_results_cp_asvlevel_univar) 

# Extract relevant columns for the heatmap
np_heatmap_data_univar <- np_results_cp_asvlevel_univar %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
np_heatmap_data_univar$metadata <- factor(np_heatmap_data_univar$metadata, levels = custom_metadata_order)

# Rename metadata variables
np_heatmap_data_univar <- np_heatmap_data_univar %>%
  mutate(metadata = dplyr::recode(metadata,
    "ecoli_RF_preds" = "RF-modelled E. coli exposure",
    "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
    "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
    "meca_RF_preds" = "RF-modelled mecA ARG exposure",
    "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
    "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
    "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
    "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))


# Create the heatmap
np_heatmap_univar <- ggplot(np_heatmap_data_univar, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (NP niche) \nUnivariable animal species")

ggsave("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/Heatmap_Maaslin_resutls_cp_np_asvlevel_univar.svg", np_heatmap_univar, width = 6, height = 8)
```
#### NP (multivariable species)
```{r}
# Define the directory
np_folder <- "../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/"

# Read NP results
np_files_cp_asvlevel <- list.files(np_folder, pattern = "np_", full.names = TRUE)
np_results_cp_asvlevel <- lapply(np_files_cp_asvlevel, read.csv)
np_results_cp_asvlevel <- bind_rows(np_results_cp_asvlevel) 

# Extract relevant columns for the heatmap
np_heatmap_data <- np_results_cp_asvlevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

np_heatmap_data$metadata
# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
np_heatmap_data$metadata <- factor(np_heatmap_data$metadata, levels = custom_metadata_order)

# Rename metadata variables
np_heatmap_data <- np_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
    "ecoli_RF_preds" = "RF-modelled E. coli exposure",
    "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
    "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
    "meca_RF_preds" = "RF-modelled mecA ARG exposure",
    "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
    "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
    "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
    "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))


# Create the heatmap
heatmap_cp_np_asvlevel <- ggplot(np_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (NP niche) \nMultivariable animal species")

ggsave("../Output/DA_analysis/MaAsLin/CP_NP/ASV_level/Heatmap_Maaslin_results_cp_np_asvlevel.svg", heatmap_cp_np_asvlevel, width = 6, height = 8)
```
#### OP (univariable species)
```{r}
# Define the directory
op_folder <- "../Output/DA_analysis/MaAsLin/CP_OP/"

# Read OP results
op_files_cp_asvlevel <- list.files(op_folder, pattern = "maaslin_cp_OP_asvlevel", full.names = TRUE)
op_results_cp_asvlevel <- lapply(op_files_cp_asvlevel, read.csv)
op_results_cp_asvlevel <- bind_rows(op_results_cp_asvlevel) 

# Extract relevant columns for the heatmap
op_heatmap_data <- op_results_cp_asvlevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
op_heatmap_data$metadata <- factor(op_heatmap_data$metadata, levels = custom_metadata_order)

# Rename metadata variables
op_heatmap_data <- op_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
    "ecoli_RF_preds" = "RF-modelled E. coli exposure",
    "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
    "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
    "meca_RF_preds" = "RF-modelled mecA ARG exposure",
    "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
    "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
    "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
    "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))

# Create the heatmap
heatmap_cp_op_asvlevel_univar <- ggplot(op_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
    ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (OP niche)\nUnivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/Heatmap_Maaslin_resutls_cp_op_asvlevel_univar.svg", heatmap_cp_op_asvlevel_univar, width = 6, height = 8)
```


#### OP (multivariable species)
```{r}
# Define the directory
op_folder <- "../Output/DA_analysis/MaAsLin/CP_OP/ASV_level"

# Read OP results
op_files_cp_asvlevel <- list.files(op_folder, pattern = "op_", full.names = TRUE)
op_results_cp_asvlevel <- lapply(op_files_cp_asvlevel, read.csv)
op_results_cp_asvlevel <- bind_rows(op_results_cp_asvlevel) 

# Extract relevant columns for the heatmap
op_heatmap_data <- op_results_cp_asvlevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
op_heatmap_data$metadata <- factor(op_heatmap_data$metadata, levels = custom_metadata_order)

# Rename metadata variables
op_heatmap_data <- op_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
    "ecoli_RF_preds" = "RF-modelled E. coli exposure",
    "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
    "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
    "meca_RF_preds" = "RF-modelled mecA ARG exposure",
    "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
    "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
    "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
    "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))

# Create the heatmap
heatmap_cp_op_asvlevel <- ggplot(op_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
    ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (OP niche)\nMultivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_OP/ASV_level/Heatmap_Maaslin_results_cp_op_asvlevel_multivar.svg", heatmap_cp_op_asvlevel, width = 6, height = 8)
```

### Genus level 
#### NP (univariable livestock species)
```{r}
# Define the directory
np_folder_genus_univar <- "../Output/DA_analysis/MaAsLin/CP_NP/"

# Read NP results
np_files_cp_genuslevel <- list.files(np_folder_genus_univar, pattern = "maaslin_cp_np_genuslevel_", full.names = TRUE)
np_results_cp_genuslevel <- lapply(np_files_cp_genuslevel, read.csv)
np_results_cp_genuslevel <- bind_rows(np_results_cp_genuslevel) 

# Extract relevant columns for the heatmap
np_heatmap_data_univar <- np_results_cp_genuslevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
np_heatmap_data_univar$metadata <- factor(np_heatmap_data_univar$metadata, levels = custom_metadata_order)

# Rename metadata variables
np_heatmap_data_univar <- np_heatmap_data_univar %>%
  mutate(metadata = dplyr::recode(metadata,
                           "ecoli_RF_preds" = "RF-modelled E. coli exposure",
                           "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
                           "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
                           "meca_RF_preds" = "RF-modelled mecA ARG exposure",
                           "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
                           "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
                           "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
                           "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))


# Create the heatmap
heatmap_cp_np_genuslevel_univar <- ggplot(np_heatmap_data_univar, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (NP niche)\nUnivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_NP/Genus_level/Heatmap_Maaslin_results_cp_np_genuslevel_univar.svg", heatmap_cp_np_genuslevel_univar, width = 5, height = 6)
```
#### NP (multivariable livestock species)
```{r}
# Define the directory
np_folder_genus <- "../Output/DA_analysis/MaAsLin/CP_NP/Genus_level/"

# Read NP results
np_files_cp_genuslevel <- list.files(np_folder_genus, pattern = "np_", full.names = TRUE)
np_results_cp_genuslevel <- lapply(np_files_cp_genuslevel, read.csv)
np_results_cp_genuslevel <- bind_rows(np_results_cp_genuslevel) 

# Extract relevant columns for the heatmap
np_heatmap_data <- np_results_cp_genuslevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
np_heatmap_data$metadata <- factor(np_heatmap_data$metadata, levels = custom_metadata_order)

# Rename metadata variables
np_heatmap_data <- np_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
                           "ecoli_RF_preds" = "RF-modelled E. coli exposure",
                           "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
                           "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
                           "meca_RF_preds" = "RF-modelled mecA ARG exposure",
                           "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
                           "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
                           "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
                           "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))


# Create the heatmap
heatmap_cp_np_genuslevel <- ggplot(np_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (NP niche)\nMultivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_NP/Genus_level/Heatmap_Maaslin_resutls_cp_np_genuslevel.svg", heatmap_cp_np_genuslevel, width=5, height=6)
```
#### OP (univariable livestock species)
```{r}
# Define the directory
op_folder_genus_univar <- "../Output/DA_analysis/MaAsLin/CP_OP/"

# Read OP results
op_files_cp_genuslevel_univar <- list.files(op_folder_genus_univar, pattern = "op_", full.names = TRUE)
op_results_cp_genuslevel_univar <- lapply(op_files_cp_genuslevel_univar, read.csv)
op_results_cp_genuslevel_univar <- bind_rows(op_results_cp_genuslevel_univar) 

# Extract relevant columns for the heatmap
op_heatmap_data_univar <- op_results_cp_genuslevel_univar %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
op_heatmap_data_univar$metadata <- factor(op_heatmap_data_univar$metadata, levels = custom_metadata_order)

# Rename metadata variables
op_heatmap_data_univar <- op_heatmap_data_univar %>%
  mutate(metadata = dplyr::recode(metadata,
                           "ecoli_RF_preds" = "RF-modelled E. coli exposure",
                           "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
                           "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
                           "meca_RF_preds" = "RF-modelled mecA ARG exposure",
                           "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
                           "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
                           "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
                           "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))

# Create the heatmap
heatmap_cp_op_genuslevel_univar <- ggplot(op_heatmap_data_univar, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (OP niche)\nUnivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_OP/Genus_level/Heatmap_Maaslin_results_cp_op_genuslevel_univar.svg", heatmap_cp_op_genuslevel_univar, width = 6, height =7)

```

#### OP (multivariable livestock species)
```{r}
# Define the directory
op_folder_genus <- "../Output/DA_analysis/MaAsLin/CP_OP/Genus_level/"

# Read OP results
op_files_cp_genuslevel <- list.files(op_folder_genus, pattern = "op_", full.names = TRUE)
op_results_cp_genuslevel <- lapply(op_files_cp_genuslevel, read.csv)
op_results_cp_genuslevel <- bind_rows(op_results_cp_genuslevel) 

# Extract relevant columns for the heatmap
op_heatmap_data <- op_results_cp_genuslevel %>%
  select(feature, metadata, coef, label) %>%
  group_by(feature, metadata) 

# Define the custom order of metadata
custom_metadata_order <- c("ecoli_RF_preds", "staph_RF_preds", "tetw_RF_preds", 
                           "meca_RF_preds", "KRD_nPigsWghtDist.3000m.sum", 
                           "KRD_nGoatsWghtDist.3000m.sum", "KRD_nPoultryWghtDist.3000m.sum", 
                           "KRD_nCowsWghtDist.3000m.sum")

# Reorder metadata variables 
op_heatmap_data$metadata <- factor(op_heatmap_data$metadata, levels = custom_metadata_order)

# Rename metadata variables
op_heatmap_data <- op_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
                           "ecoli_RF_preds" = "RF-modelled E. coli exposure",
                           "staph_RF_preds" = "RF-modelled Staphylococcus spp. exposure",
                           "tetw_RF_preds" = "RF-modelled tetW ARG exposure",
                           "meca_RF_preds" = "RF-modelled mecA ARG exposure",
                           "KRD_nPigsWghtDist.3000m.sum" = "Distance weighted number of pigs in 3000m",
                           "KRD_nGoatsWghtDist.3000m.sum" = "Distance weighted number of goats in 3000m",
                           "KRD_nPoultryWghtDist.3000m.sum" = "Distance weighted number of chickens in 3000m",
                           "KRD_nCowsWghtDist.3000m.sum" = "Distance weighted number of cattle in 3000m"
  ))

# Create the heatmap
heatmap_cp_op_genuslevel <- ggplot(op_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of Coefficients by residential\n livestock exposure (OP niche)\nMultivariable livestock species")

ggsave("../Output/DA_analysis/MaAsLin/CP_OP/Genus_level/Heatmap_Maaslin_results_cp_op_genuslevel_multivar.svg", heatmap_cp_op_genuslevel,width = 6, height =7)
```
## CP and GF
### ASV level 
#### NP 
```{r}
# Extract relevant columns for the heatmap
np_cp_gf_heatmap_data <- maaslin_cp_gf_np_ord %>%
  select(feature, metadata, coef, label)

# Rename metadata variables
np_cp_gf_heatmap_data <- np_cp_gf_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
                           "population" = "Population (GF vs CP)"))


# Create the heatmap
heatmap_cp_gf_np_asvlevel <- ggplot(np_cp_gf_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of coefficients by occupational\n livestock exposure (NP niche) - ASV level")

ggsave("../Output/DA_analysis/MaAsLin/CP_GF_NP/ASV_level/Heatmap_Maaslin_results_cp_gf_np_asvlevel.svg", heatmap_cp_gf_np_asvlevel, width = 5, height=8)
```
#### OP
```{r}
# Extract relevant columns for the heatmap
op_cp_gf_heatmap_data <- maaslin_cp_gf_op_ord %>%
  select(feature, metadata, coef, label)

# Rename metadata variables
op_cp_gf_heatmap_data <- op_cp_gf_heatmap_data %>%
  mutate(metadata = dplyr::recode(metadata,
                           "population" = "Population (GF vs CP)"))


# Create the heatmap
heatmap_cp_gf_op_asvlevel <- ggplot(op_cp_gf_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of coefficients by occupational\n livestock exposure (OP niche) - ASV level")

ggsave("../Output/DA_analysis/MaAsLin/CP_GF_OP/ASV_level/Heatmap_Maaslin_results_cp_gf_op_asvlevel.svg", heatmap_cp_gf_op_asvlevel, width = 5, height=8)
```

### Genus level 
#### NP 
```{r}
# Extract relevant columns for the heatmap
np_cp_gf_genus_heatmap_data <- maaslin_cp_gf_genus_np_ord %>%
  select(feature, metadata, coef, label)

# Rename metadata variables
np_cp_gf_genus_heatmap_data <- np_cp_gf_genus_heatmap_data %>%
  mutate(metadata = recode(metadata,
                           "population" = "Population (GF vs CP)"))


# Create the heatmap
heatmap_cp_gf_np_genuslevel <- ggplot(np_cp_gf_genus_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of coefficients by occupational\n livestock exposure (NP niche) - Genus level")

ggsave("../Output/DA_analysis/MaAsLin/CP_GF_NP/Genus_level/Heatmap_Maaslin_results_cp_gf_np_genuslevel.svg", heatmap_cp_gf_np_genuslevel, width = 5, height=8)
```
#### OP
```{r}
# Extract relevant columns for the heatmap
op_cp_gf_genus_heatmap_data <- maaslin_cp_gf_genus_op_ord %>%
  select(feature, metadata, coef, label)

# Rename metadata variables
op_cp_gf_genus_heatmap_data <- op_cp_gf_genus_heatmap_data %>%
  mutate(metadata = recode(metadata,
                           "population" = "Population (GF vs CP)"))


# Create the heatmap
heatmap_cp_gf_op_genuslevel <- ggplot(op_cp_gf_genus_heatmap_data, aes(x = metadata, y = feature, fill = coef)) +
  geom_tile(color = "white") +  # Create the heatmap tiles
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1], 
    mid = carto_pal(7, "Geyser")[4], 
    high = carto_pal(7, "Geyser")[7], 
    midpoint = 0,
    name = "Coefficient"
  ) +
  geom_text(aes(label = label), color = "white", size = 3) +  # Add labels inside the cells
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),   # Remove y-axis title
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  labs(title = "Heatmap of coefficients by occupational\n livestock exposure (op niche) - Genus level")

ggsave("../Output/DA_analysis/MaAsLin/CP_GF_op/Genus_level/Heatmap_Maaslin_results_cp_gf_op_genuslevel.svg", heatmap_cp_gf_op_genuslevel, width = 5, height=8)
```
# Check some sequences of interest using BLAST
```{r}
# Make a dataframe of all sequences in the microbiome
seq <- refseq(ps_new_metadata_RFpreds) %>%
  as.data.frame()

# ASVs of interest
which(rownames(seq) == "Moraxella_5") # 5
which(rownames(seq) == "Staphylococcus_2") # 2
which(rownames(seq) == "Corynebacterium_7") # 7
which(rownames(seq) == "Sphingomonas_175") # 131
which(rownames(seq) == "Corynebacterium_stationis_251") # 

# Check the sequences for all of these ASVs
print(seq[5, ])
print(seq[2, ])
print(seq[7, ])
print(seq[194, ])

# Extract the tax_table from the phyloseq object
tax_table_data <- tax_table(ps_new_metadata_RFpreds)
tax_table_data[1:20,]
# View the filtered data
corynebacterium_data

```


# ANCOM-BC2 analysis is not possible with a continuous variable
